---
title: "Advanced R for Econometricians"
subtitle: "Databases using R"
author: 
  - "Martin Arnold"
  - "Alexander Gerber"
date: '`r Sys.Date()`'
output:
  xaringan::moon_reader:
    seal: false
    self: true
    lib_dir: libs
    css: ["default", "../assets/sydney-fonts.css", "../assets/sydney.css", "../assets/custom.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
---
```{r, include = FALSE, cache = FALSE}
library(knitr)
knitr::opts_chunk$set(eval = TRUE)

default_source_hook <- knit_hooks$get("source")
knit_hooks$set(
  source = function(x, options) {
    if(is.null(options$table))
      default_source_hook(x, options)
    else {
      eval(parse(text = x)) %>%
        kable("html") %>%
        kable_styling("hover", full_width = F)
    }
  }
)
```


---
## Memory Limitations of R

- The amount of data you can work with in R is mainly constraint by your computers memory. 
- In times where data sets with several 100 GB or even TB are become more and more common
  loading all data into memory is usually not feasable. 
- Often only a subset or an aggregation of the data is required for the analysis which requires much less memory space.
- How to subset or aggregate data? &rarr; Use a database!

---
## (Relational) Databases
- A database is a place to store, manipulate and retreat data.
- A relational database aims to reduce redundancies by spreading the data over multiple tables and
  relate them to each other. 

#### Example of Redundent Data

.font90[
```{r, echo = FALSE, eval = TRUE, results = 'asis', message = FALSE}
library(dplyr)
set.seed(123)
cust_names<- c("17" = "Peter", "34" = "Kelly", "108" = "Aleta")
cust_address<- c("17" = "Carnaby  St.", "34" = "Downing Rd.", "108" = "Berwick St.")
sells <- tibble(store_id = rep(1:2, each = 4), 
                    store_location   = rep(c("Baker St.", "Abbey Road"), each = 4),
                    store_mngr   = rep(c("John", "George"), each = 4),
                    cust_id      = sample(c(34, 17, 108, 17, 108, 17, 34, 17), 8, replace = TRUE),
                    cust_address = cust_address[as.character(cust_id)],
                    cust_name    = cust_names[as.character(cust_id)],
                    date = sample(seq(as.Date("2019/1/1"), as.Date("2019/3/1"), 
                               by = "day"), 8),
                    total = sample(10:20, 8, replace = TRUE)
                )

knitr::kable(sells, format = "html")
```
]
---
#### Example of Non-Redundent Data 

- The same information would be stored in a relational database like this: 

```{r, eval = TRUE, echo = FALSE}
store <- tibble(store_id =  1:2, 
                store_location = c("Baker St.", "Abbey Road"),
                store_mngr   = c("John", "George")
                )
customer <- tibble(cust_id = names(cust_names),
                   cust_names = cust_names,
                   cust_address = cust_address)

sells_rel <- sells %>% select(-store_location, -store_mngr, -cust_address, -cust_name) 
```

.pull-left[
```{r, echo = FALSE, eval = TRUE}
knitr::kable(sells_rel, format = "html", caption = "Sell Records")
```
]
.pull-right[
```{r, echo = FALSE, eval = TRUE}
knitr::kable(store, format = "html", caption = "Store")
```
<br>
```{r, echo = FALSE, eval = TRUE}
knitr::kable(customer, format = "html", caption = "Customer")
```
]

<br>
- The relationship between the tables is defined by the `id` columns (called keys).



---
## R and Databases

- `dplyr` also works with remote on-disk data stored in relational databases.
- For this to work you need 
    ```{r, eval = FALSE}
install.packages("dbplyr")
    ```
- You’ll also need to install a `DBI` backend package. The `DBI` package provides a common interface that allows `dplyr` to work with many different databases using the same code. 

- Some commonly used backends are:
    - `RMySQL` connects to MySQL and MariaDB.
    - `RPostgreSQL` connects to Postgres and Redshift.
    - `RSQLite` embeds a SQLite database.
    - `bigrquery` connects to Google’s BigQuery.

---
## Connect to a Database

- To connect to a database authentication information such as 
  username and password are required (the database admin should provide those). 
- You also need to specify a driver that makes communication with the database possible.
- There is a remote MySQL database hosted on the web which we have access to:
    ```{r, eval = FALSE}
library(dplyr)
con <- DBI::dbConnect(
               drv = RMariaDB::MariaDB(), # MariaDB driver works for MySQL as well
               username = "sql7308621",
               dbname   = "sql7308621", 
               host     = "sql7.freesqldatabase.com",
               password = rstudioapi::askForPassword("Database password"),
               port     = "3306"
               )
    ```

```{r, eval = TRUE, echo = FALSE}
library(dplyr)
con <- DBI::dbConnect(
               drv = RMariaDB::MariaDB(), # MariaDB driver works for MySQL as well
               username = "sql7308621",
               dbname   = "sql7308621", 
               host     = "sql7.freesqldatabase.com",
               password = "MH1M64RSY4",
               port     = "3306"
               )
```

```{r, echo = FALSE, eval = FALSE}
DBI::dbWriteTable(con, "gapminder", gapminder::gapminder)
DBI::dbWriteTable(con, "country_codes", gapminder::country_codes)
```

---
## Access Data

- When the connection is established we can use the object `con` to communicate with the database.
- The function `DBI::dbListTables()` tells us which tables are in the database 
```{r}
DBI::dbListTables(con)
```
- With `dplyr::tbl()` we can create an object which is a reference to a table in the database. Most `dplyr` functions work on this object as it was as a data frame (or `tibble`).
```{r}
gap_db <- tbl(con, "gapminder")
class(gap_db)
```

---
## Use Standard `dplyr`  Functions on the Database

```{r, message=FALSE, warning=FALSE}
gap_avg_db <- gap_db %>% 
                group_by(continent) %>% 
                select_if(is.numeric) %>% 
                select(-year) %>%
                summarise_all(mean)

```

- Even if it seems as if `gap_avg_db` is a data frame it still is only a set of instructions to be performed by the data base. 
- `show_query()` tells us which instructions are sent to the database. 
```{r, warning=FALSE}
show_query(gap_avg_db)
```


---
## Use Standard `dplyr`  functions on the Database


When working with databases, dplyr tries to be as lazy as possible:

- It never pulls data into R unless you explicitly ask for it.
- It delays doing any work until the last possible moment: it collects everything you want to do and then sends it to the database in one step.

`collect()` actually pulls the data into R.  

```{r}
gap_avg <- gap_avg_db %>% collect()
print(gap_avg)
```

---
## Disconnect

- After the connection to the database is no longer needed you should disconnect. 

```{r}
DBI::dbDisconnect(con)
rm(con)
```



---
## Set up an SQLite Database

- SQLite is very light weighted Database Management System 
- An SQLite database is contained in a single file. There is no need for a server. 
- Everything you need is installed if the package `RSQLite` is installed. 

#### Create an SQLite Database
- A new SQLite database is automatically created if you call `DBI::dbConnect()`
with argument `dbname = "path_to_database"` where `"path_to_database"` is a path to 
a non-existing file. 
- In case `"path_to_database"` is an existing SQLite database it 
connects to that database. 

```{r, echo = TRUE, eval = FALSE}
library(dplyr)
con <- DBI::dbConnect(
               drv = RSQLite::SQLite(),
               dbname = here::here("databases","gapminder.sqlite3"), 
               )
```

---
## Write Data to the Database

- With 
    ```{r, eval = FALSE}
DBI::dbWriteTable(con, 
                  name = "mtcars_db", 
                  value = mtcars)
    ```
we add the data frame `mtcars` as a new table with name `mtcars_db` to the database.

- To add data to an existing table we use the additional argument `append = TRUE`.
    ```{r, eval = FALSE}
new_car <- data.frame(mpg = 1, cyl = 16, disp = 800, hp = 210, 
                      drat = 5, wt = 6, qsec = 2, vs = 1, am = 0)
rownames(new_car) <- "Super Car 500"

DBI::dbWriteTable(con, 
                  name = "mtcars_db", 
                  value = new_car, 
                  append = TRUE)
    ```





