---
title: "Introduction"
author: 
  - "Martin Arnold"
  - "Alexander Gerber"
date: '`r Sys.Date()`'
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["default", "assets/sydney-fonts.css", "assets/sydney.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
     #ratio: "16:9"
    #includes:
     # after_body: ../xaringan_files/terminal_highlight.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Outline
1. Introduction 
2. Prerequisites
  - Rmarkdown
  - Git and Github
3. R for Data Science 
  - Data Visualisation 
  - Data Handling
  - Modeling 
4. R Programming
  - Internals
  - Functional Programming
  - Object Oriented Programming 
5. Performance
  - Profiling and Benchmarking 
  - Parallelisation
  - Rcpp
6. Further subjects
  - web scraping 
  - shiny 
  - text mining 
  - package development 
---
class: top, left

### Getting Help

If you want to find out how a package works, search for

  - a vignitte by the package authors 
    ```{r, echo = TRUE, eval = FALSE}
    library(broom)
    vignette("broom")
    ```
  - a blog post about the package, e.g. on [R Bloggers](https://www.r-bloggers.com)
  
If you have specific questions about your code or an error messages search
  - [StackOverflow](https://stackoverflow.com/)  
  - [RStudio Community](https://community.rstudio.com/) (if it is about RStudio or packages developed by RStudio).
  - the [GitHub](https://github.com/) repository.

Probably somebody else has already asked the same questions. 

If you don't find an answer, you can ask a question yourself, what is harder then it sounds.  
Most importantly you have to provide a [reproducible example](https://stackoverflow.com/questions/5963269/how-to-make-a-great-r-reproducible-example).



---
class: top, left

