---
title: "Graphics in R"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,  warning=FALSE, message=FALSE)
```

## Data Visualization 

From Wikipedia:

"R is a programming language and free software environment for statistical computing and **graphics** ... ." 

There are three major graphical systems: 

- `R base graphics` 
- `lattice`
- `ggplot2`

In this course we will focus on `ggplot2`. If you want to learn about the other two, 
[R Graphics](https://www.crcpress.com/R-Graphics/Murrell/p/book/9780429196119)
is a great source. 

## ggplot2
What is **ggplot2**?

- an R package for data vizualisation 
- implementation of the [Grammer of Graphics](https://www.springer.com/de/book/9781475731002) in R
- part of the tidyverse

Some interesting links: 

- [R for Data Science](http://r4ds.had.co.nz/) 
- [ggplot2: Elegant graphics for data Analysis](http://moderngraphics11.pbworks.com/f/ggplot2-Book09hWickham.pdf)
- [Cookbook](http://www.cookbook-r.com/Graphs/)
- [R Graph Gallery](https://www.r-graph-gallery.com/)
- [DataCamp course](https://www.datacamp.com/courses/data-visualization-with-ggplot2-1)


## The grammer of graphics

- data: in ggplot2, data must be stored as a data.frame
- aesthetics: describe visual characteristics 
  - e.g., position (x and y), size, color, shape, fill, …
- geoms: describe type of geometric objects
  - e.g., points, lines, polygons, …
- stats: describe statistical transformations that summarize the data 
  - e.g., counts, regression lines, …
- scales: for each aesthetic, describe how visual characteristic is converted to display values 
  - e.g., x and y scales, color scales, size scales, shape scales, …
- facets: describe how the data is split into subsets and displayed as multiple small graphs
- coords: describes 2-D space that the data is projected onto 
  - e.g., Cartesian coordinates, polar coordinates, map projections, …

## A basic example

```{r, echo = TRUE, eval = FALSE}
library(ggplot2)
data("diamonds")
ggplot(data = diamonds) + geom_point(mapping = aes(x = carat, y = price))
```

`ggplot()`

- creates a coordinate system that you can add layers to

`geom_point()`

- adds a layer of points
- each geom function takes a mapping argument, which paired with `aes()`
defines how variables in your dataset are mapped to visual properties

## A basic example

```{r, echo = FALSE, eval = TRUE}
library(ggplot2)
data("diamonds")
diamonds <- diamonds[sample(1:nrow(diamonds), 5000), ]
ggplot(data = diamonds) + geom_point(mapping = aes(x = carat, y = price))
```

## Adding a layers

- To make the pattern in the data more visible we add a smoothing function to the plot 

```{r, echo = TRUE, eval = FALSE}
ggplot(data = diamonds) + 
  geom_point(mapping = aes(x = carat, y = price)) +
  geom_smooth(mapping = aes(x = carat, y = price), method = 'loess')
```

- to write more compact code we can leave out the parameter names and switch `mapping` to 
`ggplot()`
- the same mapping is then used for all layers (but can be overwritten if necessary)

```{r, echo = TRUE, eval = FALSE}
ggplot(diamonds, aes(carat, price)) + 
  geom_point() +
  geom_smooth(method = 'loess')
```

## Adding a layers

```{r, echo = FALSE, eval = TRUE}
ggplot(diamonds, aes(x = carat, y = price)) + 
  geom_point() +
  geom_smooth(method = 'loess')
```

## Aesthetics

- until now we only used the x and y coordinates as aesthetics
- `?geom_point()` tells us about further aesthetics that we can map data to 
- note that each geom has its own set of aesthetics

```{r, echo = TRUE, eval = FALSE}
ggplot(diamonds, aes(x = carat, y = price, color = color, shape =  cut)) + 
  geom_point()
```

## Aesthetics

```{r, echo = FALSE, eval = TRUE, warning=FALSE}
ggplot(diamonds, aes(x = carat, y = price, color = color, shape =  cut)) + 
  geom_point()
```

## Statitic Transformations and Aesthetics

- if we map aesthetics to discrete variables, then ggplot will automatically group the data accordingly
- in this case every statistic transformation is performed by group
- you can see what this means by the following example

```{r, echo = TRUE, eval = FALSE, warning=FALSE}
ggplot(diamonds, aes(x = carat, y = price, color = color, shape =  cut)) + 
  geom_point() + 
  geom_smooth()
```

## Statitic Transformations and Aesthetics
```{r, echo = FALSE, eval = TRUE, warning=FALSE}
ggplot(diamonds, aes(x = carat, y = price, color = color, shape =  cut)) + 
  geom_point() +
  geom_smooth()
```

## Statitic Transformations and Aesthetics
- If this is not desired simply move color and shape as additional aesthetics to `geom_point()`
```{r, echo = TRUE, eval = FALSE}
ggplot(diamonds, aes(x = carat, y = price)) + 
  geom_point(aes(color = color, shape = cut)) +
  geom_smooth()
```

- in some circumstances it might also be convinient to leave all aesthetics in `ggplot()` (e.g. if there are many layers) and overwrite the created groups by a arbitrary constant value

```{r, echo = TRUE, eval = FALSE}
ggplot(diamonds, aes(x = carat, y = price, color = color, shape = cut)) + 
  geom_point() +  
  geom_smooth(aes(group = 1))
```


## Statistic Transformations and Aesthetics
```{r, echo = FALSE, eval = TRUE}
ggplot(diamonds, aes(x = carat, y = price, color = color, shape = cut)) + 
  geom_point() +
  stat_smooth(aes(group = 1))
```

## Geometric Objects ans Statistic transformations

- we can define a new layer in terms of a geometric object (`geom_...`) or a statistic transformation (`stat_...`)
- each geometric object is associated with a default statistic transformation and vice verse
- e.g. `geom_point()` has the indentity function as statistic transformation and `geom_smooth()` fits a regression model before plotting a line with a prediction interval
- there is is an alias function `stat_smooth()` which does esentially the same, however, in the first case we could change the statistic transformation and in the second case we could change the geometric object
- most of the time it is not I good idea to change the default behaviour (e.g. try  `geom_point(stat = "smooth", method = "lm")`) but we will see an example where it can be useful

## Bar plot
`geom_bar()` counts the number of observations with within each group
```{r, echo = TRUE, eval = TRUE}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut)) # x should be discreate
```

## Histogram 

```{r, echo = TRUE, eval = TRUE}
ggplot(diamonds, aes(x = depth, fill = cut)) + 
  geom_histogram(binwidth = 0.1) 
```

## 1D Density 

```{r, echo = TRUE, eval = FALSE}
ggplot(diamonds, aes(x = depth, fill = cut)) + # try color instead of fill 
  geom_density() 
```

```{r, echo = FALSE, eval = TRUE}
ggplot(diamonds, aes(x = depth, fill = cut)) + # try color instead of fill 
  geom_density() 
```

## 2D Density

```{r, echo = TRUE, eval = FALSE}
ggplot(data = diamonds, mapping = aes(carat, price)) + 
  geom_point() + 
  geom_density2d()
```

```{r, echo = FALSE, eval = TRUE}
ggplot(data = diamonds, mapping = aes(carat, price)) + 
  geom_point() + 
  geom_density2d()
```

## 2D Density with geom = "polygon"

```{r echo = TRUE, eval = FALSE}
ggplot(data = diamonds, mapping = aes(carat, price)) + 
  geom_point() + 
  stat_density2d(aes(fill = ..level..), geom = "polygon")
```

```{r, echo = FALSE, eval = TRUE}
ggplot(data = diamonds, mapping = aes(carat, price)) + 
  geom_point() + 
  stat_density2d(aes(fill = ..level..), geom = "polygon")

```


## Faceting
- Faceting generates small multiples each showing a different subset of the data
- helps to compare patterns in different parts of the data and see whether they are the same or different

```{r, echo = TRUE, eval = FALSE}
ggplot(data = diamonds, mapping = aes(carat, price))  + 
  geom_point() + 
   stat_density2d(aes(fill = ..level..), geom = "polygon") + 
  facet_grid( ~ cut)
```

```{r, echo = TRUE, eval = FALSE}
ggplot(data = diamonds, mapping = aes(carat, price))  + 
  geom_point() + 
   stat_density2d(aes(fill = ..level..), geom = "polygon") + 
  facet_grid(color ~ cut)
```

## Faceting using one variable
```{r, echo = FALSE, eval = TRUE}
ggplot(data = diamonds, mapping = aes(carat, price))  + 
  geom_point() + 
   stat_density2d(aes(fill = ..level..), geom = "polygon") + 
  facet_grid( ~ cut)
```

## Faceting using two variables
```{r, echo = FALSE, eval = TRUE}
ggplot(data = diamonds, mapping = aes(carat, price))  + 
  geom_point() + 
   stat_density2d(aes(fill = ..level..), geom = "polygon") + 
  facet_grid(color  ~ cut)
```

## Position adjustments
- each geometric object has a parmeter responsible for position adjustment
- in the following example we devided the bar plot into groups defined by the variable clarity
- we can stack the bars of each group on top of each other
```{r, echo = TRUE, eval = FALSE}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "stack") # the default
```
- we can put the bars for each group next to each other 
```{r, echo = TRUE, eval = FALSE}
  geom_bar(., position = "dodge") 
```
- we can compare relative proportions at each x by stacking the bars and  standardising each bar to have the same height

```{r, echo = TRUE, eval = FALSE}
geom_bar(., position = "dodge") 
```

## Stacked bar plot 
```{r, echo = FALSE, eval = TRUE}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "stack") # the default
```

## Dodged bar plot 
```{r, echo = FALSE, eval = TRUE}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "dodge") 
```

## Filled bar plot

```{r, echo = FALSE, eval = TRUE}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "fill") 
```

## Scales

- scales determine how the date are mapped to the asthetics (e.g. which value takes which color)
- scales are defined by functions of the form `scale_aestheticname_scalename()`

```{r, echo = TRUE, eval = TRUE}
ggplot(diamonds, aes(carat, price, color = cut)) + geom_point() + scale_color_grey()
```


## Scales

- you can also use your own scales

```{r, echo = TRUE, eval = TRUE}
ggplot(diamonds, aes(carat, price, color = cut)) + 
  geom_point() + 
  scale_color_manual(values = c("#c9792e", "blue", "green", "gray", "thistle2"))
```


## Titles and labels 

With the function `labs()` you can 
- add a title and a subtible
- add a caption
- add a tag
- change the axis lables
- change the legend tilte

```{r, echo = TRUE, eval = TRUE}
ggplot(diamonds, aes(x = carat, y = price, fill = cut)) + 
  geom_point() +
  labs(title = "Diamanten", x = "Karat", y = "Preis", fill = "Schnitt") 
```

## Titles and labels 
```{r, echo = FALSE, eval = TRUE}
ggplot(diamonds, aes(x = carat, y = price, fill = cut)) + 
  geom_point() +
  labs(title = "Diamanten", x = "Karat", y = "Preis", fill = "Schnitt") 
```


## Themes

Themes control the appearance of the plot 

- fontype and fontsize
- background
- ticks marks and labels
- grid lines
- ...

There are many predefined themes, however, if you like you can define all by yourself


## Themes

```{r}
ggplot(diamonds, aes(carat, price, color = cut)) + 
  geom_point() + theme_bw()
```



