---
title: "Advanced R for Econometricians"
subtitle: "Data Visualization with ggplot2"
author: "Alexander Gerber"
date: "5 November 2018"
output:
  xaringan::moon_reader:
    css: ["default", "../assets/sydney-fonts.css", "../assets/sydney.css", "../assets/title_slides.css"]
    self_contained: false # if true, fonts will be stored locally
    seal: false # show a title slide with YAML information
    includes:
      in_header: "../assets/mathjax-equation-numbers.html"
    nature:
      beforeInit: ["../assets/remark-zoom.js", "https://platform.twitter.com/widgets.js"]
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9' # alternatives '16:9' or '4:3' or others e.g. 13:9
      navigation:
        scroll: false # disable slide transitions by scrolling
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE, 
                      fig.align = 'center', 
                      dpi = 250, fig.width=10, out.width='72%')
options(htmltools.dir.version = FALSE)
```
class: title-slide title-ggplot2 center middle

# `r rmarkdown::metadata$title`
## `r rmarkdown::metadata$subtitle`

---
class: left, top
## Data Visualization 

From Wikipedia:

"R is a programming language and free software environment for statistical computing and **graphics** ... ." 

There are three major graphical systems: 

- R base graphics
- `lattice`
- `ggplot2`

In this course, we will focus on `ggplot2`. If you want to learn about the others, 
[R Graphics](https://www.crcpress.com/R-Graphics/Murrell/p/book/9780429196119)
is a great source. 

---
class: left, top

## ggplot2
What is **ggplot2**?

- an R package for data visualisation 
- implementation of the [Grammar of Graphics](https://www.springer.com/de/book/9781475731002) in R
- part of the [tidyverse](https://www.tidyverse.org/)

Some interesting links: 

- [R for Data Science](http://r4ds.had.co.nz/) 
- [ggplot2: Elegant graphics for data Analysis](http://moderngraphics11.pbworks.com/f/ggplot2-Book09hWickham.pdf)
- [R Graphics Cookbook](http://www.cookbook-r.com/Graphs/)
- [R Graph Gallery](https://www.r-graph-gallery.com/)
- [DataCamp](https://www.datacamp.com/courses/data-visualization-with-ggplot2-1)

---
class: left, top
## The Layered Grammar of Graphics

The grammar consists of 
- data
- aesthetic mappings (e.g. mapping of data to x and y coordinates, size, color, shape, ...)
- geometric objects (e.g. points, lines, bars, ...)
- scales (controls mapping from data to aesthetics ,e.g., which colors should be used)
- facets (splitting data to create plots for subgroups)
- statistical transformations (summarize data before plotting)
- coordinate systems (e.g. Cartesian, semi-log, polar, ...)

The data, mappings, statistical transformations and geometric objects form a layer.
A plot can have multiple layers. 

---
class: left, top
## A Basic Example

```{r, echo = TRUE, eval = FALSE}
library(ggplot2)
data("diamonds")
ggplot(data = diamonds) + 
  geom_point(mapping = aes(x = carat, y = price))
```

`ggplot()`

- Creates a coordinate system that you can add layers to.
- Everything you provide as an argument will be the default for all added layers.

`geom_point()`

- Adds a layer of points.
- Each `geom_*` function takes a mapping argument, which paired with `aes()`
defines how variables in your data are mapped to visual properties.

---
class: left, top
## A Basic Example

```{r, echo = FALSE, eval = TRUE}
library(ggplot2)
set.seed(321)
data("diamonds")
diamonds <- diamonds[sample(1:nrow(diamonds), 5000), ]
ggplot(data = diamonds) + geom_point(mapping = aes(x = carat, y = price))
```

---
class: left, top
## Adding Layers

To make the (possibly nonlinear) relationship in the data easier visible we add a smoothing function to the plot.

```{r, echo = TRUE, eval = FALSE}
ggplot(data = diamonds) + 
  geom_point(mapping = aes(x = carat, y = price)) +
  geom_smooth(mapping = aes(x = carat, y = price), method = 'loess')
```

To write more compact code we can
- omit the parameter names 
- switch `mapping` to `ggplot()`.    

  
The same mapping is then used for all layers (but can be overwritten if necessary).

```{r, echo = TRUE, eval = FALSE}
ggplot(diamonds, aes(carat, price)) + 
  geom_point() +
  geom_smooth(method = 'loess')
```

---
class: left, top
## Adding a Layer

```{r, echo = FALSE, eval = TRUE}
ggplot(diamonds, aes(x = carat, y = price)) + 
  geom_point() +
  geom_smooth(method = 'loess')
```

---
class: left, top
## Aesthetics

- Until now we only used the x and y coordinates as aesthetics.
- `?geom_point()` tells us about further aesthetics that we can map data to. 
- Each geom has its own set of aesthetics.

```{r, echo = TRUE, eval = FALSE}
ggplot(diamonds, aes(x = carat, y = price,
                     color = color, shape =  cut)) + 
  geom_point()
```

---
class: left, top
## Aesthetics

```{r, echo = FALSE, eval = TRUE}
ggplot(diamonds, aes(x = carat, y = price, 
                     color = color, shape =  cut)) + 
  geom_point()
```

---
## Exercise

1. Experiment with `geom_point()` using the `mtcars` data set. Try out different aesthetics with different variables. What do you note?  

???
Specifically explain the different behaviour of factor and numeric variables for e.g. color. 



---
class: left, top
## Statistical  Transformations and Aesthetics

- If discrete variables are mapped to aesthetics, ggplot will automatically group the data.
- In this case every statistic transformation is performed by group.


```{r, echo = TRUE, eval = FALSE, warning=FALSE}
ggplot(diamonds, aes(x = carat, y = price, 
                     color = color, shape =  cut)) + 
  geom_point() + 
  geom_smooth()
```

---
class: left, top
## Statistical Transformations and Aesthetics
```{r, echo = FALSE, eval = TRUE, fig.height=6}
ggplot(diamonds, aes(x = carat, y = price, color = color, shape =  cut)) + 
  geom_point() +
  geom_smooth()
```

---
class: left, top
## Statistical Transformations and Aesthetics
- If this is not desired simply move color and shape as additional aesthetics to `geom_point()`.
    ```{r, echo = TRUE, eval = FALSE}
ggplot(diamonds, aes(x = carat, y = price)) + 
  geom_point(aes(color = color, shape = cut)) +
  geom_smooth()
    ```
- Sometimes it is more convenient to leave all aesthetics in `ggplot()` (e.g. if there are many layers) and overwrite the created groups by an arbitrary constant value.
    ```{r, echo = TRUE, eval = FALSE}
ggplot(diamonds, aes(x = carat, y = price, 
                     color = color, shape = cut)) + 
  geom_point() +  
  geom_smooth(aes(group = 1))
    ```

---
class: left, top
## Statistical Transformations and Aesthetics
```{r, echo = FALSE, eval = TRUE}
ggplot(diamonds, aes(x = carat, y = price, color = color, shape = cut)) + 
  geom_point() +
  stat_smooth(aes(group = 1))
```

---
class: left, top
## Geometric Objects and Statistic Transformations

- Layers may be defined in terms of a geometric object (`geom_*`) or a statistical transformation (`stat_*`).
- Each geometric object is associated with a default statistical transformation and vice versa.
     **Examples**: 

    - `geom_point()` has the identity function as statistical transformation.

    - `geom_smooth()` fits a regression model before plotting a line with a prediction interval.
    
- `stat_smooth()` is an alias which does essentially the same. However, in the first case we could change the statistic transformation and in the second case we could change the geometric object.
- Often it is not a good idea to change the default behaviour (e.g. try  `geom_point(stat = "smooth", method = "lm")`) but we will see an example where it can be useful.

---
### Exercises
1. Use the `mtcars` data set and plot `mpg` vs. `hp`. Add a smoothing line to the plot. 
2. Add for each number of cylinders a linear smoothing function to the plot.
3. Find out how to remove the confidence interval. 
4. Instead of a simple linear model use a quatratic function for smoothing.  
 



---
class: left, top
## Bar Plot
`geom_bar()` counts the number of observations within each group and produces a bar plot. 
```{r, echo = TRUE, eval = TRUE, out.width='50%'}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut)) # x should be discrete
```

---
class: left, top
## Histogram 

```{r, echo = TRUE, eval = TRUE, out.width='50%'}
ggplot(diamonds, aes(x = depth, fill = cut)) + 
  geom_histogram(binwidth = 0.1) 
```

---
class: left, top
## 1D Density 

```{r, echo = TRUE, eval = FALSE, out.width='50%'}
ggplot(diamonds, aes(x = depth, fill = cut)) + # try color instead of fill 
  geom_density() 
```

```{r, echo = FALSE, eval = TRUE, out.width='50%'}
ggplot(diamonds, aes(x = depth, fill = cut)) + # try color instead of fill 
  geom_density() 
```

---
class: left, top
## 2D Density

```{r, echo = TRUE, eval = FALSE}
ggplot(data = diamonds, mapping = aes(carat, price)) + 
  geom_point() + 
  geom_density2d()
```

```{r, echo = FALSE, eval = TRUE, out.width='55%'}
ggplot(data = diamonds, mapping = aes(carat, price)) + 
  geom_point() + 
  geom_density2d()
```

---
class: left, top
## 2D Density with `geom = "polygon"`

```{r echo = TRUE, eval = FALSE}
ggplot(data = diamonds, mapping = aes(carat, price)) + 
  geom_point() + 
  stat_density2d(aes(fill = ..level..), geom = "polygon")
```

```{r, echo = FALSE, eval = TRUE, out.width='55%'}
ggplot(data = diamonds, mapping = aes(carat, price)) + 
  geom_point() + 
  stat_density2d(aes(fill = ..level..), geom = "polygon")

```

---
class: left, top
## Faceting
- Faceting generates multiple plots each showing a different subset of the data.

```{r, echo = TRUE, eval = FALSE}
ggplot(data = diamonds, mapping = aes(carat, price))  + 
  geom_point() + 
  facet_grid( ~ cut)
```

---
class: left, top
## Faceting using one variable
```{r, echo = FALSE, eval = TRUE}
ggplot(data = diamonds, mapping = aes(carat, price))  + 
  geom_point() + 
  facet_grid( ~ cut)
```

---
class: left, top
## Faceting using two variables
```{r, echo = FALSE, eval = TRUE}
ggplot(data = diamonds, mapping = aes(carat, price))  + 
  geom_point() + 
  facet_grid(color  ~ cut)
```

---
class: left, top
## Position Adjustments
- Each geometric object has a parameter for position adjustment.
    ```{r, echo = TRUE, eval = FALSE}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "stack") # the default
    ```
- Instead of stacking the bars we can position them side-by-side using `dodge`. 
    ```{r, echo = TRUE, eval = FALSE}
  geom_bar(., position = "dodge") 
    ```
- With `fill` relative proportians can be compared.
    ```{r, echo = TRUE, eval = FALSE}
geom_bar(., position = "fill") 
    ```

---
class: left, top
## Stacked Bar Plot 
```{r, echo = FALSE, eval = TRUE}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "stack") # the default
```

---
class: left, top
## Dodged Bar Plot 
```{r, echo = FALSE, eval = TRUE}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "dodge") 
```

---
class: left, top
## Filled Bar Plot
```{r, echo = FALSE, eval = TRUE}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "fill") 
```

---
class: left, top
## Scales

- Scales determine how the data are mapped to the aesthetics (e.g. which value takes which color).
- Scales are defined by functions of the form `scale_aestheticname_scalename()`.

```{r, echo = TRUE, eval = TRUE, out.width='50%'}
ggplot(diamonds, aes(carat, price, color = cut)) + geom_point() + scale_color_grey()
```

---
class: left, top
## Scales
- You can also define your own scales.

```{r, echo = TRUE, eval = TRUE, out.width='50%'}
ggplot(diamonds, aes(carat, price, color = cut)) + 
  geom_point() + 
  scale_color_manual(values = c("#c9792e", "blue", "green", "gray", "thistle2"))
```

---
class: left, top
## Titles and labels 

`labs()` allows you to
- add a title and a subtitle
- add a caption
- add a tag
- change the axis labels
- change the legend title.

```{r, echo = TRUE, eval = FALSE}
ggplot(diamonds, aes(x = carat, y = price, fill = cut)) + 
  geom_point() +
  labs(title = "Diamanten", x = "Karat", y = "Preis", fill = "Schnitt") 
```

---
class: left, top
## Titles and labels 
```{r, echo = FALSE, eval = TRUE}
ggplot(diamonds, aes(x = carat, y = price, fill = cut)) + 
  geom_point() +
  labs(title = "Diamanten", x = "Karat", y = "Preis", fill = "Schnitt") 
```

---
class: left, top
## Themes

Themes control the appearance of the plot 

- font type and font size
- background
- ticks marks and labels
- grid lines
- ...

There are many predefined themes. However, if you like you may also define everything yourself.

```{r, echo = TRUE, eval = FALSE}
ggplot(diamonds, aes(carat, price, color = cut)) + 
  geom_point() + theme_bw()
```

---
class: left, top
## Themes

```{r}
ggplot(diamonds, aes(carat, price, color = cut)) + 
  geom_point() + theme_bw()
```

---
class: left, top

## Exercises 

1. Download the Titanic data set from Moodle and 

- use a bar plot to show how many people survived the Titanic compared to those who didn't. 
- add a color coding to the previous plot to visualize the differences between the passangers gender.
- split the previous plot into three plots based on the passangers class. 
- compare the age distribution between the survivors and non-survivors.  

<ol start="2">
<li> Try to answer the following questions about the <TT>mpg</TT> dataset (comes with <TT>ggplot2</TT>) using <TT>ggplot</TT>.</li>
</ol>
<ul>
    <li> How are engine size and fuel economy related?</li>
    <li> Do certain manufacturers care more about economy than others?</li>
    <li> Has fuel economy improved in the last ten years?</li>
</ul>

<ol start="3">
<li> Compare the two data sets  <TT>economics</TT> and <TT>economics_long</TT> (both come with ggplot) with respect to the ease of use when working with <TT>ggplot</TT>.</li>

---
<ol start="4">
<li>Repreoduce the following plot with <TT>ggplot</TT>.</li>

```{r, echo = TRUE, eval = FALSE}
plot(mtcars$mpg ~ mtcars$wt, xlab = "wt", ylab = "mpg", pch = 19, ylim = c(5, 35))

mod <- lm(mpg ~ wt, data = mtcars)
abline(mod, col = "red")

wt_new <- seq(min(mtcars$wt), max(mtcars$wt), by = 0.05)

conf_interval <- predict(mod, newdata = data.frame(wt = wt_new), 
                         interval = "confidence", level = 0.95)

# setup vertrices of polygon (for shading the CI):
p <- cbind(c(wt_new, rev(wt_new)), c(conf_interval[, 3], rev(conf_interval[, 2])))

polygon(p, col = adjustcolor("steelblue", alpha.f = 0.5), )

lines(wt_new, conf_interval[, 2], col = "steelblue", lty = 2)
lines(wt_new, conf_interval[, 3], col = "steelblue", lty = 2)
```






```{r, eval = FALSE}
library(ggplot2)
titanic <- read.csv(here::here("datasets", "titanic.csv"), dec = ",")

write.csv(titanic, file = here::here("datasets", "titanic.csv"))

ggplot(titanic) + geom_bar(aes(x = survived, fill = sex), position = "dodge")
ggplot(titanic) + geom_bar(aes(x = survived, fill = sex), position = "dodge") + facet_wrap(~pclass)
ggplot(titanic) + geom_bar(aes(x = survived, fill = sex), position = "dodge") + facet_wrap( ~age < 15)
ggplot(titanic) + geom_density(aes(x = age, col = survived), position = "dodge")  + facet_grid( pclass~sex)
ggplot(titanic) + geom_density(aes(x = age, col = survived), position = "dodge")  + facet_grid( pclass~sex)
ggplot(titanic,aes(x = fare, y = age, col = survived)) + geom_point(position = "dodge")  + facet_grid( pclass~sex) + geom_smooth()
```


```{r, eval = FALSE}
ggplot(mpg, aes(y = hwy, x  = displ, col = factor(year))) + geom_point() + geom_smooth()
```


```{r, eval = FALSE}
economics_long
ggplot(economics_long, aes(x = date, y = value, col = variable)) + geom_line()
```

