---
title: "Data Visualization with ggplot2"
author: "Alexander Gerber"
date: "5 November 2018"
output:
  xaringan::moon_reader:
    css: ["default", "../assets/sydney-fonts.css", "../assets/sydney.css"]
    self_contained: false # if true, fonts will be stored locally
    seal: true # show a title slide with YAML information
    includes:
      in_header: "../assets/mathjax-equation-numbers.html"
    nature:
      beforeInit: ["../assets/remark-zoom.js", "https://platform.twitter.com/widgets.js"]
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9' # alternatives '16:9' or '4:3' or others e.g. 13:9
      navigation:
        scroll: false # disable slide transitions by scrolling
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE)
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include=FALSE}
# # solarized_dark(
# #   code_font_family = "Fira Code",
# #   code_font_url    = "https://cdn.rawgit.com/tonsky/FiraCode/1.204/distr/fira_code.css"
# # )
# library(xaringanthemer)
# 
# mono_accent(
#     base_color           = "#09017F",
#     header_font_google   = google_font("Roboto", "700"),
#     text_font_google     = google_font("Roboto Condensed"),
#     code_highlight_color = "#D2B6E8",
#     code_font_family     = "Fira Code",
#     code_font_url        = "https://cdn.rawgit.com/tonsky/FiraCode/1.204/distr/fira_code.css"
#     )

```

class: left, top
## Data Visualization 

From Wikipedia:

"R is a programming language and free software environment for statistical computing and **graphics** ... ." 

There are three major graphical systems: 

- R base graphics
- `lattice`
- `ggplot2`

In this course, we will focus on `ggplot2`. If you want to learn about the other two, 
[R Graphics](https://www.crcpress.com/R-Graphics/Murrell/p/book/9780429196119)
is a great source. 

---
class: left, top

## ggplot2
What is **ggplot2**?

- an R package for data visualisation 
- implementation of the [Grammar of Graphics](https://www.springer.com/de/book/9781475731002) in R
- part of the tidyverse

Some interesting links: 

- [R for Data Science](http://r4ds.had.co.nz/) 
- [ggplot2: Elegant graphics for data Analysis](http://moderngraphics11.pbworks.com/f/ggplot2-Book09hWickham.pdf)
- [Cookbook](http://www.cookbook-r.com/Graphs/)
- [R Graph Gallery](https://www.r-graph-gallery.com/)
- [DataCamp](https://www.datacamp.com/courses/data-visualization-with-ggplot2-1)

---
class: left, top
## The Layered Grammer of Graphics

The grammar consists of 
- data
- aesthetic mappings (e.g. mapping of data to x and y coordinates, size, color, shape, ...)
- geometric objects (e.g. points, lines, bars, ...)
- scales (controls mapping from data to aesthetics  e.g. which colors to use )
- facets (splitting data to create plots for subgroups)
- statistical transformation (summarize data before plotting)
- coordinate systems (e.g. Cartesian, semi-log, polar, ...)

The data, mappings, statistical transformation and geometric objects form a layer.
A plot can have multiple layers. 

---
class: left, top
## A Basic Example

```{r, echo = TRUE, eval = FALSE}
library(ggplot2)
data("diamonds")
ggplot(data = diamonds) + 
  geom_point(mapping = aes(x = carat, y = price))
```

`ggplot()`

- creates a coordinate system that you can add layers to
- takes as arguments the default for all added layers

`geom_point()`

- adds a layer of points
- each geom function takes a mapping argument, which paired with `aes()`
defines how variables in your dataset are mapped to visual properties

---
class: left, top
## A Basic Example

```{r, echo = FALSE, eval = TRUE, fig.width=10}
library(ggplot2)
set.seed(321)
data("diamonds")
diamonds <- diamonds[sample(1:nrow(diamonds), 5000), ]
ggplot(data = diamonds) + geom_point(mapping = aes(x = carat, y = price))
```

---
class: left, top
## Adding Layers

To make the pattern in the data better visible we add a smoothing function to the plot.

```{r, echo = TRUE, eval = FALSE}
ggplot(data = diamonds) + 
  geom_point(mapping = aes(x = carat, y = price)) +
  geom_smooth(mapping = aes(x = carat, y = price), method = 'loess')
```

To write more compact code we can
- leave out the parameter names 
- switch `mapping` to `ggplot()`.    

  
The same mapping is then used for all layers (but can be overwritten if necessary).

```{r, echo = TRUE, eval = FALSE}
ggplot(diamonds, aes(carat, price)) + 
  geom_point() +
  geom_smooth(method = 'loess')
```

---
class: left, top
## Adding a Layer

```{r, echo = FALSE, eval = TRUE, fig.width=10}
ggplot(diamonds, aes(x = carat, y = price)) + 
  geom_point() +
  geom_smooth(method = 'loess')
```

---
class: left, top
## Aesthetics

- Until now we only used the x and y coordinates as aesthetics.
- `?geom_point()` tells us about further aesthetics that we can map data to. 
- Each geom has its own set of aesthetics.

```{r, echo = TRUE, eval = FALSE}
ggplot(diamonds, aes(x = carat, y = price,
                     color = color, shape =  cut)) + 
  geom_point()
```

---
class: left, top
## Aesthetics

```{r, echo = FALSE, eval = TRUE, fig.width=10}
ggplot(diamonds, aes(x = carat, y = price, 
                     color = color, shape =  cut)) + 
  geom_point()
```

---
class: left, top
## Statistic  Transformations and Aesthetics

- If discrete variables are mapped to aesthetics, then ggplot will automatically group the data.
- In this case every statistic transformation is performed by group.


```{r, echo = TRUE, eval = FALSE, warning=FALSE}
ggplot(diamonds, aes(x = carat, y = price, 
                     color = color, shape =  cut)) + 
  geom_point() + 
  geom_smooth()
```

---
class: left, top
## Statitic Transformations and Aesthetics
```{r, echo = FALSE, eval = TRUE, fig.width=10, fig.height=6}
ggplot(diamonds, aes(x = carat, y = price, color = color, shape =  cut)) + 
  geom_point() +
  geom_smooth()
```

---
class: left, top
## Statitic Transformations and Aesthetics
- If this is not desired simply move color and shape as additional aesthetics to `geom_point()`
```{r, echo = TRUE, eval = FALSE}
ggplot(diamonds, aes(x = carat, y = price)) + 
  geom_point(aes(color = color, shape = cut)) +
  geom_smooth()
```

- in some circumstances, it might also be convenient to leave all aesthetics in `ggplot()` (e.g. if there are many layers) and overwrite the created groups by an arbitrary constant value

```{r, echo = TRUE, eval = FALSE}
ggplot(diamonds, aes(x = carat, y = price, 
                     color = color, shape = cut)) + 
  geom_point() +  
  geom_smooth(aes(group = 1))
```

---
class: left, top
## Statistic Transformations and Aesthetics
```{r, echo = FALSE, eval = TRUE, fig.width=10, fig.height=6}
ggplot(diamonds, aes(x = carat, y = price, color = color, shape = cut)) + 
  geom_point() +
  stat_smooth(aes(group = 1))
```

---
class: left, top
## Geometric Objects and Statistic transformations

- We can define a new layer in terms of a geometric object (`geom_...`) or a statistic transformation (`stat_...`).
- Each geometric object is associated with a default statistic transformation and vice verse.
- E.g. `geom_point()` has the identity function as statistic transformation and `geom_smooth()` fits a regression model before plotting a line with a prediction interval.
- There is is an alias function `stat_smooth()` which does essentially the same, however, in the first case we could change the statistic transformation and in the second case we could change the geometric object.
- Most of the time it is not a good idea to change the default behaviour (e.g. try  `geom_point(stat = "smooth", method = "lm")`) but we will see an example where it can be useful.

---
class: left, top
## Bar Plot
`geom_bar()` counts the number of observations with within each group
```{r, echo = TRUE, eval = TRUE, fig.width=10, fig.height=5}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut)) # x should be discreate
```

---
class: left, top
## Histogram 

```{r, echo = TRUE, eval = TRUE, fig.width=10, fig.height=5}
ggplot(diamonds, aes(x = depth, fill = cut)) + 
  geom_histogram(binwidth = 0.1) 
```

---
class: left, top
## 1D Density 

```{r, echo = TRUE, eval = FALSE, fig.width=10, fig.height=5}
ggplot(diamonds, aes(x = depth, fill = cut)) + # try color instead of fill 
  geom_density() 
```

```{r, echo = FALSE, eval = TRUE, fig.width=10, fig.height=5}
ggplot(diamonds, aes(x = depth, fill = cut)) + # try color instead of fill 
  geom_density() 
```

---
class: left, top
## 2D Density

```{r, echo = TRUE, eval = FALSE}
ggplot(data = diamonds, mapping = aes(carat, price)) + 
  geom_point() + 
  geom_density2d()
```

```{r, echo = FALSE, eval = TRUE, fig.width=10, fig.height=5}
ggplot(data = diamonds, mapping = aes(carat, price)) + 
  geom_point() + 
  geom_density2d()
```

---
class: left, top
## 2D Density with geom = "polygon"

```{r echo = TRUE, eval = FALSE, fig.width=10, fig.height=5}
ggplot(data = diamonds, mapping = aes(carat, price)) + 
  geom_point() + 
  stat_density2d(aes(fill = ..level..), geom = "polygon")
```

```{r, echo = FALSE, eval = TRUE, fig.width=10, fig.height=5}
ggplot(data = diamonds, mapping = aes(carat, price)) + 
  geom_point() + 
  stat_density2d(aes(fill = ..level..), geom = "polygon")

```

---
class: left, top
## Faceting
- Faceting generates multiple plots each showing a different subset of the data.
- Helps to compare patterns in different parts of the data.

```{r, echo = TRUE, eval = FALSE, fig.width=10}
ggplot(data = diamonds, mapping = aes(carat, price))  + 
  geom_point() + 
   stat_density2d(aes(fill = ..level..), geom = "polygon") + 
  facet_grid( ~ cut)
```

---
class: left, top
## Faceting using one variable
```{r, echo = FALSE, eval = TRUE, fig.width=10}
ggplot(data = diamonds, mapping = aes(carat, price))  + 
  geom_point() + 
   stat_density2d(aes(fill = ..level..), geom = "polygon") + 
  facet_grid( ~ cut)
```

---
class: left, top
## Faceting using two variables
```{r, echo = FALSE, eval = TRUE, fig.width=10}
ggplot(data = diamonds, mapping = aes(carat, price))  + 
  geom_point() + 
   stat_density2d(aes(fill = ..level..), geom = "polygon") + 
  facet_grid(color  ~ cut)
```

---
class: left, top
## Position adjustments
- each geometric object has a parameter responsible for position adjustment
- in the following example, we divided  the bar plot into groups defined by the variable clarity
- we can stack the bars of each group on top of each other
```{r, echo = TRUE, eval = FALSE, fig.width=10}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "stack") # the default
```
- we can put the bars for each group next to each other 
```{r, echo = TRUE, eval = FALSE, fig.width=10}
  geom_bar(., position = "dodge") 
```
- we can compare relative proportions at each x by stacking the bars and  standardising each bar to have the same height

```{r, echo = TRUE, eval = FALSE, fig.width=10}
geom_bar(., position = "dodge") 
```

---
class: left, top
## Stacked bar plot 
```{r, echo = FALSE, eval = TRUE, fig.width=10}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "stack") # the default
```

---
class: left, top
## Dodged bar plot 
```{r, echo = FALSE, eval = TRUE, fig.width=10}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "dodge") 
```

---
class: left, top
## Filled bar plot
```{r, echo = FALSE, eval = TRUE, fig.width=10}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "fill") 
```

---
class: left, top
## Scales

- scales determine how the data are mapped to the aesthetics (e.g. which value takes which color)
- scales are defined by functions of the form `scale_aestheticname_scalename()`

```{r, echo = TRUE, eval = TRUE, fig.width=10, fig.height=4}
ggplot(diamonds, aes(carat, price, color = cut)) + geom_point() + scale_color_grey()
```

---
class: left, top
## Scales
- you can also use your own scales

```{r, echo = TRUE, eval = TRUE, fig.width=10, fig.height=4}
ggplot(diamonds, aes(carat, price, color = cut)) + 
  geom_point() + 
  scale_color_manual(values = c("#c9792e", "blue", "green", "gray", "thistle2"))
```

---
class: left, top
## Titles and labels 

With the function `labs()` you can 
- add a title and a subtitle
- add a caption
- add a tag
- change the axis labels
- change the legend title

```{r, echo = TRUE, eval = FALSE}
ggplot(diamonds, aes(x = carat, y = price, fill = cut)) + 
  geom_point() +
  labs(title = "Diamanten", x = "Karat", y = "Preis", fill = "Schnitt") 
```

---
class: left, top
## Titles and labels 
```{r, echo = FALSE, eval = TRUE, fig.width=10}
ggplot(diamonds, aes(x = carat, y = price, fill = cut)) + 
  geom_point() +
  labs(title = "Diamanten", x = "Karat", y = "Preis", fill = "Schnitt") 
```

---
class: left, top
## Themes

Themes control the appearance of the plot 

- font type and font size
- background
- ticks marks and labels
- grid lines
- ...

There are many predefined themes, however, if you like you can define everything by yourself.

```{r, echo = TRUE, eval = FALSE}
ggplot(diamonds, aes(carat, price, color = cut)) + 
  geom_point() + theme_bw()
```

---
class: left, top
## Themes

```{r, fig.width=10}
ggplot(diamonds, aes(carat, price, color = cut)) + 
  geom_point() + theme_bw()
```

---
class: left, top

## Exercises 

1. Follow the instructions about the Titanic data set (download from Moodle).

- Use a bar plot to show how many people survived the Titanic compared to those who didn't. 
- Add a color coding to the previous plot to visualize the diffences between the passangers gender.
- Split the previous plot into three plots based on the passangers class. 
- Compare the age distribution between the survivors and non-survivors.  

<ol start="2">
<li> Try to answer the following questions about the <TT>mpg</TT> dataset (comes with <TT>ggplot2</TT>) using <TT>ggplot</TT>.  </li>
</ol>
<ul>
    <li> How are engine size and fuel economy related?</li>
    <li> Do certain manufacturers care more about economy than others?</li>
    <li> Has fuel economy improved in the last ten years?</li>
</ul>

<ol start="3">
<li> Compare the two data sets  <TT>economics </TT> and <TT>economics_long </TT> (both come with ggplot) with respect to the ease of use when working with <TT>ggplot </TT>.</li>

```{r, eval = FALSE}
library(ggplot2)
titanic <- read.csv(here::here("datasets", "titanic.csv"), dec = ",")

write.csv(titanic, file = here::here("datasets", "titanic.csv"))

ggplot(titanic) + geom_bar(aes(x = survived, fill = sex), position = "dodge")
ggplot(titanic) + geom_bar(aes(x = survived, fill = sex), position = "dodge") + facet_wrap(~pclass)
ggplot(titanic) + geom_bar(aes(x = survived, fill = sex), position = "dodge") + facet_wrap( ~age < 15)
ggplot(titanic) + geom_density(aes(x = age, col = survived), position = "dodge")  + facet_grid( pclass~sex)
ggplot(titanic) + geom_density(aes(x = age, col = survived), position = "dodge")  + facet_grid( pclass~sex)
ggplot(titanic,aes(x = fare, y = age, col = survived)) + geom_point(position = "dodge")  + facet_grid( pclass~sex) + geom_smooth()
```


```{r, eval = FALSE}
ggplot(mpg, aes(y = hwy, x  = displ, col = factor(year))) + geom_point() + geom_smooth()
```


```{r, eval = FALSE}
economics_long
ggplot(economics_long, aes(x = date, y = value, col = variable)) + geom_line()
```

