---
title: "Advanced R for Econometricians"
subtitle: "Data Wrangling with `dplyr` and `tidyr`"
author: "Alexander Gerber"
date: "5 November 2018"
output:
  xaringan::moon_reader:
    css: ["default", "../assets/sydney-fonts.css", "../assets/sydney.css"]
    self_contained: true # if true, fonts will be stored locally
    seal: false # show a title slide with YAML information
    includes:
      in_header: "../assets/mathjax-equation-numbers.html"
    nature:
      beforeInit: ["../assets/remark-zoom.js", "https://platform.twitter.com/widgets.js"]
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9' # alternatives '16:9' or '4:3' or others e.g. 13:9
      navigation:
        scroll: false # disable slide transitions by scrolling
---
```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE)
```
class: title-slide center middle

# `r rmarkdown::metadata$title`
## `r rmarkdown::metadata$subtitle`

---
class: top, left
## Tibble
- Tibbles are the main data structure used in the tidyverse and can be viewed as a modern version of the data frame. 

- `tibble()` creates a new tibble from scratch.
    ```{r, echo = TRUE, eval = TRUE}
    library(tidyverse)
    new_tibble <- tibble(x = 1:3, y = letters[1:3])
    ```

- `as_tibble()` creates a tibble from an existing object (e.g. a data frame or a matrix).
    ```{r, echo = TRUE, eval = TRUE}
    a_data.frame <- data.frame(x = 1:3, y = letters[1:3])
    a_tibble     <- as_tibble(a_data.frame)
    ```

- Everything that works for data frames also works with tibbles because
    ```{r, echo = TRUE, eval = TRUE}
    class(a_tibble)
    ```



---
class: top, left

## Tibble vs. data.frame
- Why should we use tibbles instead of data frames? See the following comparison.  
    ```{r, echo = TRUE}
data(mtcars) 
mtcars_tibble <- as_tibble(mtcars)
    ```

- different print method
    ```{r, echo = TRUE, eval = FALSE}
    print(mtcars)
    print(mtcars_tibble)
    ```

- more consistent subsetting behaviour
    ```{r, echo = TRUE, eval = FALSE}
mtcars[ ,1:2] # Return a data.frame
mtcars[ ,1]   # Returns a vector 
mtcars_tibble[ ,1:2] # Returns a tibble
mtcars_tibble[ ,1]   # Returns a tibble
    ```

---
## Tibble vs. data.frame
- lazy and sequential evaluation 
    ```{r, echo = TRUE, eval = FALSE}
data.frame(x = 1:3, y = x^2) # not working
tibble(x = 1:3, y = x^2)
    ```

- does not change input types
    ```{r, echo = TRUE, results='hold'}
text <- letters[1:3]
class(text)
df   <- data.frame(text) # stringsAsFactors = TRUE is the default 
class(df$text)
tib  <- tibble(text)
class(tib$text)
    ```

- For more see `vignette("tibble")`.

---
class: top, left
## dplyr 

The `dplyr` package is the most important package of the `tidyverse` when it comes to data manipulation and transformation. 

The core characteristics of the package:

- focus on data frames (data frame in &rarr; data frame out)
- verbose compared to base R which leads to longer but easier to read code
- consistent syntax. 

You can compare the functionality to that of a  relational database which provides methods to work with
single tables and multiple tables. 


---
class: top, left
## Single Table Functions

First, we look at functions that work on a single table. 
 
- `filter()` to select rows based on their values.
    ```{r, echo = TRUE, eval = FALSE}
filter(mtcars_tibble, hp >= 100 & mpg >= 20)
    ```

- `slice()` to select rows by position. 
    ```{r, eval = FALSE, echo = TRUE}
slice(mtcars_tibble, 10:15)
    ```
- `select()`/`rename()` to select/rename variables based on their names.
    ```{r, echo = TRUE, eval = FALSE}
select(mtcars_tibble, cyl, hp)
    ```

---
class: top, left
## Single Table Functions

- `arrange()` to reorder rows.
    ```{r, echo = TRUE, eval = FALSE}
arrange(mtcars_tibble, cyl)
arrange(mtcars_tibble, desc(cyl))
    ```

- `mutate()` and `transmute()` to add new variables that are functions of existing variables. 
    ```{r, echo = TRUE, eval = FALSE}
    mutate(mtcars_tibble, gpm = 1/mpg)
    transmute(mtcars_tibble, gpm = 1/mpg) 
    ```

- `summarise()` to condense multiple values to a single value.
    ```{r, echo = TRUE, eval = FALSE}
    summarise(mtcars_tibble, mean_hp = mean(hp), max_mpg = max(mpg))
    ```

---
class: top, left
## Single Table Functions
- `sample_n()` and `sample_frac()` to take random samples.
    ```{r, echo = TRUE, eval = FALSE}
    sample_n(mtcars_tibble, 10)
    sample_frac(mtcars_tibble, 0.1)
    ```

- `distinct()` removes duplicate rows.
    ```{r, echo = TRUE, eval = FALSE}
    tibble_with_duplicates    <- tibble(x = c(1,1,2,3), 
                                        y = c(4,4,5,6))
    tibble_without_duplicates <- distinct(tibble_with_duplicates)
    ```

- `count()` and `tally()` count rows.  
    ```{r, echo = TRUE, eval = FALSE}
    count(mtcars_tibble); count(mtcars_tibble, cyl)
    tally(mtcars_tibble); tally(mtcars_tibble, cyl)
    ```


---
class: top, left

## Filter 

The function `filter()` selects rows based on their values. 

- Use comparison operators `==`, `>`, `>=`, `<`, `<=`, and `!=` to filter rows based on a condition. 
    ```{r, echo = TRUE, eval = FALSE}
filter(mtcars_tibble, hp == 100)
    ```

- Use logical operators such as  `&` (and) and `|` (or) to combine multiple conditions. 
    ```{r, echo = TRUE, eval = FALSE}
filter(mtcars_tibble, hp > 100 & cyl == 4)
filter(mtcars_tibble, hp > 100 | cyl == 4)
    ```

---
- Additional arguments to `filter()` are combined with `&` which is the same as the above. 
    ```{r, echo = TRUE, eval = FALSE}
filter(mtcars_tibble, hp > 100, cyl == 4)
```

- For the common operation `x >= left & x <= right` there is the shorthand `between(x, left, right)`.
    ```{r, echo = TRUE, eval = FALSE}
filter(mtcars_tibble, x >= 90 & x <= 110)
filter(mtcars_tibble, between(hp, 90, 110)
    ```

---
## Select

The function `select()` selects columns based on their name. 

- The function can handle standard and non-standard evaluation (more about this later). 
    ```{r, echo = TRUE, eval = FALSE}
  select(mtcars_tibble, disp, vs) 
  select(mtcars_tibble, "disp", "vs") 
    ```

- You can also work with character vectors. 
    ```{r, echo = TRUE, eval = FALSE}
  select(mtcars_tibble, c("disp", "vs"))
    ```

- Use `-` to select all except the provided columns.
    ```{r, echo = TRUE, eval = FALSE}
  select(mtcars_tibble, -disp, -vs) 
  select(mtcars_tibble, -c("disp", "vs")) 
    ```

- There is a whole bunch of helper functions for `select()` on the next slides. 

---
class: top, left

## Helper Functions for `select()`

- Select columns whose name contains a character string.
    ```{r, echo=TRUE, eval = FALSE}
select(iris, contains("."))
    ```

- Select columns whose name ends with a character string.
    ```{r, echo=TRUE, eval = FALSE}
select(iris, ends_with("Length"))
    ```

- Select columns whose name matches a regular expression.
    ```{r, echo=TRUE, eval = FALSE}
select(iris, matches(".t."))
    ```

- Select columns named x1, x2, x3, x4, x5.
    ```{r, echo=TRUE, eval = FALSE}
select(iris, num_range("x", 1:5))
    ```

- Select all columns from Sepal.Width to Petal.Width.
    ```{r, echo=TRUE, eval = FALSE}
select(iris, Sepal.Width:Petal.Width)
    ```


---
class: top, left
## Helper Functions for `select()`
- Select columns whose names are in a group of names.
    ```{r, echo=TRUE, eval = FALSE}
select(iris, one_of(c("Species", "Genus")))
    ```

- Select columns whose name starts with a character string.
    ```{r, echo=TRUE, eval = FALSE}
select(iris, starts_with("Sepal"))
    ```

- Select all columns between Sepal.Length and Petal.Width (inclusive).
    ```{r, echo=TRUE, eval = FALSE}
select(iris, Sepal.Length:Petal.Width)
    ```

- Select all columns except Species. 
    ```{r, echo=TRUE, eval = FALSE}
select(iris, -Species)
    ```

- Select everything
    ```{r, echo=TRUE, eval = FALSE}
select(iris, everything())
    ```

---
## Mutate

The function `mutate()` adds new columns or replaces old columns based on existing columns. 
```{r, echo=TRUE, eval = FALSE}
mutate(mtcars_tibble,
       gpm = 1/mpg,
       disp =  disp - mean(disp) # replaces disp 
       )
```

- Use `lag()` and `lead()` if you want to access values that are not in the same row (e.g. for time series data). 
    ```{r, echo=TRUE, eval = FALSE}
    rw <- tibble(t = 1:10, y = cumsum(rnorm(10)))
    mutate(rw, increments =  y - lag(y, k = 1))
    ```

- Use `min_rank()` to add a column with the rank (`?min_rank()` gives you an overview of further ranking functions).
    ```{r, echo=TRUE, eval = FALSE}
    mutate(mtcars_tibble,
           hp_rank_smallest_first = min_rank(hp),
           hp_rank_largest_first  = min_rank(desc(hp))
       )
    ```

---
## Summarise

`summarise()` can be used to collapse columns to a single row used with e.g.: 
- base R functions such as `mean()`, `sd()`, `max()`
- dplyr specific functions for
  - position `first()`, `nth()`, `last()`
  - counts `n()`, `n_disticnt()`
  
```{r, echo = TRUE, eval = FALSE}
summarise(mtcars_tibble,
          mean_hp    = mean(hp),
          max_wt     = max(wt), 
          n          = n(),            
          third_cyl  = nth(cyl, 3) 
          )
```

- functions you have written yourself 
    ```{r, echo = TRUE, eval = FALSE}
my_variance <- function(x) 1/(n()-1) * sum((x - mean(x))^2)
summarise(by_cyl, my_variance(mpg))
    ```
---
## Summarise

- You cannot use `summarise()` with functions that return more than one value per group. 
    ```{r, eval = FALSE, echo = TRUE}
two_return_vaules <- function(x) c(mean(x), sd(x))
summarise(by_cyl, two_return_vaules(mpg))
    ```

- the function can take multiple variable or fixed inputs.
    ```{r, echo = TRUE, eval = FALSE}
two_input_vaules <- function(x, y) sum(x + y)
summarise(by_cyl, two_input_vaules(hp, qsec))       # variable inputs
summarise(by_cyl, two_input_vaules(hp, trim = 0.1)) # fixed input
    ```

---
class: top, left
## Exercises

1. Load the `nycflights13::flights` data set. Does it come formated as a tibble? 
2. Find all flights that 
  - had an arrival delay of two or more hours.
  - flew to Houston (IAH or HOU).
  - were delayed by at least an hour, but made up over 30 minutes in flight.
  - departed between midnight and 6am. 
3. Which flights travelled the longest? Which travelled the shortest?  
4. Brainstorm as many ways as possible to select `dep_time`, `dep_delay`, `arr_time`, and `arr_delay`.
5. Create a new variable `first_flight` which takes the value `TRUE` if it is the first flight of the day. 


```{r, echo=FALSE, eval = FALSE}
library(tidyverse)
library(nycflights13)
filter(flights, day == 1)


```



---
class: top, left
## Grouping

To exploit the full potential of `summarise()` we need to pair it with `group_by()`. 

```{r, echo = TRUE, results='markup'}
by_cyl <- group_by(mtcars_tibble, cyl)
print(summarise(by_cyl, mean(mpg)))
```

Use `ungroup()` remove the grouping, when you want to work on the whole data set again. 

---
class: top, left
## Grouping

It is also possible to use `group_by()` together with a lot of other single table operations.

```{r, echo = TRUE, eval = FALSE}
by_cyl <- group_by(mtcars_tibble, cyl)

# group mean is used for centering
mutate(by_cyl, centered_mpg = mpg - mean(mpg))   
transmute(by_cyl, centered_mpg = mpg - mean(mpg))

# return only first 3 rows for each group
slice(by_cyl, 1:3)  

# counts the number of rows for each row
count(by_cyl)   

# returns all groups with more than 10 rows
filter(by_cyl, n() > 10)  

# sorts one group after the other
arrange(by_cyl, disp)                          
```
---
class: top, left
## The Pipe

Usually, we combine multiple operations, e.g. first we filter, then we mutate, group and finally summarise. 
To do this we could save an intermediate result after each step like this:

```{r, echo = TRUE}
filtered   <- filter(mtcars_tibble, cyl <= 8)
mutated    <- mutate(filtered, gpm = 1/mpg)
grouped    <- group_by(mutated, gear)
summarized <- summarise(grouped, count = n(), 
                        mean_hp = mean(hp),
                        mean_gpm = mean(gpm)) 
```

This code has the disadvantage that we need to save intermediate results that we do not care about. 
---
class: top, left
## The Pipe

We could nest all those functions to get rid of the intermediate result.

```{r, echo = TRUE}
summarise(group_by(mutate(filter(mtcars_tibble, cyl <= 8), gpm = 1/mpg)), 
          count = n(), 
          mean_hp = mean(hp),
          mean_gpm = mean(gpm))
```

However, readability gets worse and debugging will become much harder. 
---
class: top, left
## The Pipe

That is where the pipe operator `%>%` from the `magrittr` comes in handy.
The value on the left side of the operator becomes the input to the function on the right side. 

```{r, echo = TRUE}
mtcars_tibble %>% 
  mutate(gpm = 1/mpg) %>%
  group_by(gear) %>% 
  summarise(count = n(), 
            mean_hp = mean(hp),
            mean_gpm = mean(gpm)) 
  
```

The pipe operator is not limited to `dplyr` or the `tidyverse`

```{r, echo = TRUE, eval = FALSE}
1:6 %>% mean
```


```{r, eval = FALSE, cache=TRUE}
library(data.table)
library(tidyverse)
crime <- fread("https://data.boston.gov/dataset/6220d948-eae2-4e4b-8723-2dc8e67722a3/resource/12cb3883-56f5-47de-afa5-3b1cf61b257b/download/tmppj4rb047.csv")
crime %>% group_by(DISTRICT) %>% summarise(N = n()) %>% arrange(desc(N)) 
crime %>% filter( OFFENSE_CODE_GROUP == "Drug Violation") %>% group_by(OFFENSE_CODE_GROUP, DISTRICT) %>% summarise(N = n()) %>% arrange(desc(N)) 


d1  <- crime %>% arrange(YEAR) %>% slice(1:10000) %>% group_by(DISTRICT, SHOOTING) %>% summarise(N = n()) 
d2  <- crime %>% arrange(YEAR) %>% sample_n(10000) %>% group_by(DISTRICT) %>% summarise(N = n()) 
plot(x = factor(d1$DISTRICT), d1$N)
points(x = factor(d2$DISTRICT), d2$N)
```

```{r, eval = FALSE}
plot(table(crime$DISTRICT[1:10000])/sum(table(crime$DISTRICT[1:10000])))
points(table(crime$DISTRICT)/sum(table(crime$DISTRICT)), col = alpha("red", 0.4), lty = 3, lwd = 5)
```


---
class: top, left
## Scoped Functions 

Often operations on all or several columns at once should be done. In this case, the standard 
functions are pedestrian. Most helpful are the scoped versions of  `summarise()` and `mutate()`.

- Use `summarise_all()`, if you want to summarise every column.
    ```{r, eval = FALSE, echo = TRUE}
mtcars_tibble %>% summarise_all(mean)
    ```

- Use `summarise_if()` to logically subset the columns to be summarised. 
    ```{r, eval = FALSE, echo = TRUE}
iris %>% summarise_if(is.numeric, mean)
    ```

-  Use `summarise_if()`, if you want to summarise some specific columns `summarise_at()`.
    ```{r, eval = FALSE, echo = TRUE}
iris %>% summarise_at(c("Sepal.Length", "Sepal.Width"), mean)
    ```



---
class: top, left
## Advanced Subjects 

If `summarise()` is to limited, because it can only handle functions that have a single 
return value one can use 
  - `group_modify()` returns a grouped tibble. In that case .f must return a data frame.
  - `group_map()` returns a list of results from calling .f on each group.
  - `group_walk()` calls .f for side effects and returns the input .tbl, invisibly.
  
Those functions are closely related to the functions in the `purrr` package which we 
will talk about later. Here we will only give a short example of how they can be used. 

```{r, echo = TRUE, eval = FALSE}
mtcars %>%
  group_by(cyl) %>% 
  group_map(~ lm(mpg  ~ hp, data = . ))

mtcars %>%
  group_by(cyl) %>% 
  group_modify(~ broom::tidy(lm(mpg  ~ hp, data = . )))

par(mfrow = c(2,2))
mtcars %>%
  group_by(cyl) %>% 
  group_walk(~ plot(mpg  ~ hp, data = . ))
```

---
class: top, left

## Exercises

Use dplyr on the Boston crime data to:

1. Find out in which district the most incident where reported. 
2. Create an ordered list of the top ten districts according to the number of drug violations. Start with the district with most drug violations. 
3. What is the most dangerous hour of the week?
4. Calculate for each district the proportion of motor vehicle accidents on all reported incidents. 
5. Sort the data set chronological by year and month. Then take the first 10.000 observations and compute the proportion of incidents where a 
shooting was involved for each district. Now take a random sample of size 10.000. Is there a noteworthy difference?


---
class: top, left
## Functions for 2 Tables

```{r, echo = TRUE, results='markup'}
band_members   
band_instruments
```
---
class: top, left
## Left Join

A left join keeps all observations of the left table
```{r, echo = TRUE, results='markup'}
band_members %>% left_join(band_instruments, by = "name")
```
---
class: top, left
## Inner Join

An inner join keeps only observations if there are matches in both tables
```{r, echo = TRUE, results='markup'}
band_members %>% inner_join(band_instruments, by = "name")
```
---
class: top, left
## Full Join

A full (outer) join keeps all observations
```{r, echo = TRUE, results='markup'}
band_members %>% inner_join(band_instruments, by = "name")
```
---
class: top, left
## Semi Join

A semi-join keeps all observations of the left table where there are matching values in the right table.
It keeps only tables from the left. 
```{r, echo = TRUE, results='markup'}
band_members %>% semi_join(band_instruments, by = "name")
```
---
class: top, left
## Anti Join

```{r, echo = TRUE, results='markup'}
band_members %>% anti_join(band_instruments, by = "name")
```

---
class: top, left
## Reshaping Data with tidy R

The goal of tidyr is to help you create tidy data. Tidy data is data where:

- Each variable is in a column.
- Each observation is a row.
- Each value is a cell.

We will talk about 

- `gather()` takes multiple columns, and gathers them into key-value pairs: it makes “wide” data longer.
- `spread()` takes two columns (key & value), and spreads into multiple columns: it makes “long” data wider.
- `separate()` takes a single column and turns it into multiple columns based on a separator.  
- `unite()` takes multiple columns and pastes them together into one column.  

---
class: top, left
## Tidy Data

```{r, echo = TRUE}
tidyr::table1
```

Each observation is in one row and each variable is one column. 

---
class: top, left
## Untidy data

```{r, echo = TRUE}
tidyr::table2
```

Observations are scattered across multiple rows


---
class: top, left
## Spreading

The function `tidyr::spread()` takes as arguments 

- `key`, containing the variable names
- `value`, containing the values for the variables in the `key` column. 

```{r, echo = TRUE}
table2 %>%
    spread(key = type, value = count)
```

---
## Untidy Data

```{r, echo = TRUE}
tidyr::table4b
```

- Columns do not cointain variables and column names are values. 

---
## Gathering

The function `gather()` takes as arguments

- the column names which are actual values
- `key`, the name of the newly created variable resulting from the column name
- `value` the name of the newly created variable.

```{r, echo = TRUE}
table4a %>% 
  gather(`1999`, `2000`, key = "year", value = "cases")
```

---
## Untidy Data

```{r, echo = TRUE}
tidyr::table3 
```

- One column contains two variables. 

---
## Separating 

The function `sparate()` can split one column into multiple columns. 
As input it takes: 
- `col`, the column to be separated
- `into`, the names of the newly created columns
- `sep`, a regular expression which tells the function where to split (by default any non-alphanumeric character ). 
- `convert`, a logical value which controls whether or not to run `run type.convert()` on the new columns.

```{r, echo = TRUE}
table3 %>% 
  separate(rate, into = c("cases", "population"), convert = TRUE)
```

---
## Untidy Data

```{r, echo = TRUE}
tidyr::table5 
```

- One variable is spread over multiple columns.  

---
## Unite

The function `unite()` pastes multiple columns together. 
As input it takes: 
- `col`, the name of the new column
- `...`. the columns to be pasted together
- `sep`, the separator between values.

```{r, echo = TRUE}
table5 %>% 
  unite(new, century, year, sep = "")
```

---
## Exercises
1. Compute the rate for `table2`, and `table4a` + `table4b`.
2. Recreate the plot given below using `table2` instead of `table1`. 
    ```{r, echo = TRUE, eval = FALSE}
library(ggplot2)
ggplot(table1, aes(year, cases)) + 
  geom_line(aes(group = country), colour = "grey50") + 
  geom_point(aes(colour = country))
    ```




