---
title: "Writing R Packages"
author: "Alexander Gerber"
date: "5 November 2018"
output:
  xaringan::moon_reader:
    css: ["default", "../assets/sydney-fonts.css", "../assets/sydney.css"]
    self_contained: false # if true, fonts will be stored locally
    seal: false # show a title slide with YAML information
    includes:
      in_header: "../assets/mathjax-equation-numbers.html"
    nature:
      beforeInit: ["../assets/remark-zoom.js", "https://platform.twitter.com/widgets.js"]
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9' # alternatives '16:9' or '4:3' or others e.g. 13:9
      navigation:
        scroll: false # disable slide transitions by scrolling
---
```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, warning=FALSE, message=FALSE)
```

http://www.math.ncu.edu.tw/~chenwc/R_note/reference/package/R-exts.pdf

---
## Minimal Package

The smallest usable R package consists of 

- an `R/` directory, containing the R Code.
- a `DESCRIPTION` file containning meta data about the package.
- a `NAMESPACE` file which defines which functions are imported and exported. 

By running

```{r, echo = TRUE, eval = FALSE}
usethis::create_package("path/to/package/pkgname")
```
a minimal package gets automatically created. It will also create and open an R project for your package. 

Since we know how to work with Git we also want to use it for package development. 
Git can be easily initialized with: 
```{r, eval = FALSE, echo = TRUE}
usethis::use_git()
```

???
```{r, eval = FALSE, echo = TRUE}
usethis::create_package(here::here("Packages", "example"))
```

---
## Write your first own package

We start by writing a first little package together. The steps are 
based on [R packages: The whole game](https://r-pkgs.org/whole-game.html). 

Before we start we should think about what our package should do. 
Since `factors` in R are often frustrating we could write a package which make working with `factors` easier. 


--
Have you ever tried to combine 2 factors?

```{r, eval = FALSE}
(a <- factor(c("character", "hits", "your", "eyeballs")))
(b <- factor(c("but", "integer", "where it", "counts")))
c(a, b)
```

--
```{r, eval = TRUE, echo = FALSE}
(a <- factor(c("character", "hits", "your", "eyeballs")))
(b <- factor(c("but", "integer", "where it", "counts")))
c(a, b)
```
--
Lets write a function called `fbind()` that does what one would expect.

---
## Where to save your function? 

All R code of the package should be in `.R` files in the `R/` directory. Put the function you have just written into the file `R/fbind.R`. 

With 
```{r, eval = FALSE}
usethis::use_r("fbind.R")
```
a file gets automatically created in the correct folder.  
If you call this function using an existing file in `R\` it will open the corresponding script. 
---
##How to test if your function works in the package environment?

Right now this might be a bit of an overkill, however, if the package becomes more complicated 
you should use `load_all()` to make all functions in the `R\` directory available. 

`load_all()` simulates the process of building, installing, and attaching the package which makes it easy to test if the function works as expected within the package. The RStudio shortcut for this is Ctrl + Shift + L.
---
## Check

After we have checked that the function is working, we also want to check if the package as a whole is working.

Run `check()` to do so. 

**Read the output of the check!** Deal with problems early and often. It’s just like incremental development of .R and .Rmd files. The longer you go between full checks that everything works, the harder it becomes to pinpoint and solve your problems.

You should get 2 warnings: 
- Non-standard license specification
- Undocumented code objects: 'fbind'

---
## Description File

The description file holds the meta data of the package. 

You can change the name and title of the package. There is also room for 
a little description. Most parts of this file are only relevant when you think 
about distribute your package e.g. via CRAN. 

However, to get rid of the licence warning use `licence: CC0`. This puts your package in the public domain.
  
Some more information about licenses can be found here: 
[License: Who can use your package?](http://r-pkgs.had.co.nz/description.html#license)

Another impotant field is `Imports:` where you have declare which other packages
you use within your package. More about this later. 


---
## Documentation

The second warning told us that a documentation for our function is missing. Again, 
if it is only for private useage this might not be too important. But if the package should be published 
on CRAN a documentation is required. 

The documentation is created when a package is build based on the  files in the folder `man/`. For 
our function `fbind()` we need a file `man/fbind.Rd`. However, we don't need to edit this file ourself.

#### roxygen2
`roxygen2` is a package that helps us with the documentation. We basically put in front of 
our function definitions something like:  

.code70[
```{r}
#' Bind two factors
#'
#' Create a new factor from two existing factors, where the new factor's levels
#' are the union of the levels of the input factors.
#'
#' @param a factor
#' @param b factor
#'
#' @return factor
#' @export
#' @examples
#' fbind(iris$Species[c(1, 51, 101)], PlantGrowth$group[c(1, 11, 21)])
```
]

---
## Documentation
A simple template for the documentation gets created by RStudio if you navigate the courser into a function call and click ***Code > Insert roxygen skeleton*** or Ctr + Alt + Shift + R.

After this is done run `devtools::document()`.
Now, there should be a `man/` directory with a file in it.

---
## NAMESPACE

The call to `devtools::document()` not only creates the documentation but also keeps the 
NAMESPACE file up-to-date. 

The export directive in NAMESPACE is what makes `fbind()` available to a user after attaching `foofactors` via `library(foofactors)`. We actually told `roxygen` to export `fbind()` by putting 
```{r}
#' @export
```

in front of the function call. 

Your NAMESPACE file should now look so
```{r}
# Generated by roxygen2: do not edit by hand
export(fbind)
```


---
## Install the package

`check()` the package again to see if the warnings are gone. Then run `install()` 
to put your package into your library. 

We can no attach the package using `library()`. 

---
## Import Packages

Often we want to use functions from other packages. To use these we first need to import 
packages from the namespace of other packages. 


Add a function that combines `lm()` and `summary()` to your package. 
Then run check. 
What die you notice?  

--
Even functions from the always available `stats` (e.g. `stats::lm()`) or `utils` (`utils::str()`) package should be imported (importent for publishing a package on CRAN). 

Declare your general intent to use some functions from the `stats` namespace with
```{r}
usethis::use_package("stats")
```

This will add the stats package to the import field of the `DESCRIPTION` file. 

If you later want to call a function from that package use `packagname::function()`. 

---
## More about Namespaces

What output do you expect for this code?
```{r}
library(dplyr)
library(MASS)

mtcars %*% select(mpg)
```
--
Do you have an idea why this isn't working?  <br>
--
<br>
The problem is that  both the `dplyr` and the `MASS` package export a function with the name
`select()`. The one in the package is used which was loaded last. 

Namespaces are a way to disambiguate which function we want to use. We can tell R to look for 
the function with the name `select()` exported by the `dplyr` package by calling `dplyr::select()`. 

---
## Imports

We already know that we need to put packages which we are using need to be specified in the 
import field of the `DESCRIPTION` file. This makes sure that the those packages are installed if our package is installed and `dplyr::select()` will work within the package. 


We can use `NAMESPACE` to define which functions from other packages we want to use without having to use`::`. 

```{r}
import(dplyr) # import all functions from one package
importFrom(dplyr, select) # import selected functions 
```

Generelly, it is better too use `::` to make it explicit where a funcion comes from.

---
## TO DO

- Put package on github
- Load package from somebody else













